Evaluating Perceptual Judgements on 3D Printed Bar Charts

### 2024-01-22 reviewer_response.qmd

---
title: "Response to Reviewers"
format: pdf
---

## Editor Suggestions

- A more extensive literature review is needed; for instance, the authors claim on page 2 that "...the situation is different now than it was in 1984...it has even changed since 2014," but there are very few citations about this. I think even mentioning some application areas (with specific papers cited) where 3D charts and printed 3D charts are increasingly prevalent is important here to motivate why readers should be interested in this study.

> Fixed. It would be hard to cite printed 3D chart prevalence as there are very few applications of this at the moment (though there are a few US maps and things on Thingiverse). We expect they will be more common moving forward, though, which is why it's important to revisit the perception of these charts.

- Given that the authors claim this is a partial replication of Cleveland and McGill's study, I suggest creating a whole section devoted to more deeply explaining their work, strengths/limitations, and how the new study differs and improves on their work. This will help particularly for those not familiar with their study.

> We have added some limited additional clarity, but have also cited papers which go through the study in more depth, including the Annual Review paper "Testing Statistical Charts: What Makes A Good Graph" by Vanderplas et al.

- In its current state, the mixed model implemented in Section 3.2 needs to be vastly simplified; as both the authors and Reviewer 1 remark, this study simply has too few participants for this parameterized model to be sufficiently fit. At the very least, I would suggest fitting a simpler model and discussing this further (including assumptions), perhaps leaving a brief description of this mixed model as something of interest in a follow-up study or in an Appendix.

> The gam model we are currently using (after following up on Reviewer 2's questions) is a bit simpler but also much more interesting. We hope this model provides more useful results than the mixed-effects model in the previous version of this paper, which was pretty clearly not appropriate given the nonlinearity in the data.

## Reviewer 1 Suggestions

- This experiment seems very far from a replication of the source material. This is somewhat softened in the paper by calling it a “partial” replication and acknowledging many of the shortcomings of the study, but even with this: the statement seems inaccurate. My reasons are as follows:
  - The participants were allowed to practice, with the true feedback given. Was there such a practice round in the original study? This seems like something that would improve later performance.

> Participants were practicing with the slider mechanism, but no feedback was provided as to the correctness of the slider position. During beta testing of the survey, we noticed that some participants were defaulting to estimating A/A+B instead of A/B, and so we added a clarifying demo page.

  - A sliding error bar is a significantly different mechanism than simply stating a ratio. The experiment in this paper is, in a sense, asks participants to translate from a visual domain to a second visual domain. If the aim of the former visualization is to get to the latter visualization, why not present the latter visualization in the first place? What information would we like someone consuming a graph to get? This paper seems to want to comment on which medium best communicates the true numbers themselves, but what is tested is whether someone is able to translate one visual medium into another visual medium. It is the authors who then translate from this latter visual medium to the ratio.

> We agree that a sliding bar is different from estimating a ratio - the latter requires both the visual assessment of the ratio as well as translating this visual assessment into a number. We had hoped to reduce cognitive load by foregoing the numeric translation step; an added benefit to this is that it is typically very difficult to model numerical estimates from participants due to rounding effects that are censored but are not censored uniformly. We do have data from a different study suggesting that estimates using sliders are similar to numerical estimates (minus the rounding effects) - this paper will be submitted within the next 6 months or so but is not quite finished. At any rate, while we agree that the method for recording data is different (and that this might be an explanation for our failure to explicitly replicate the effects in the original paper), we are also working with very different technological constraints. The charts in Cleveland & McGill had axes which could be used to help participants with estimation; our 3D printed charts do not, which is one very real limitation of that medium. Because of this issue, as well as the cognitive load and technological considerations (e.g. we have the option of a slider, in 1984 that was much less feasible), we stand by our decision to deviate from the original protocol.

  - Why are the data where a participant incorrectly identified the smaller bar removed from the study? Did they do this in the original study? Is this information irrelevant? How many observations were removed?

> The original study is not particularly specific about the methods employed. Attention check questions are fairly standard in online studies, and we have added some language clarifying the number of observations removed. Fundamentally, if participants don't identify the smaller bar, it is not clear what the slider ratio would even represent, as it is fundamentally confined to $[0,1]$ and the participant isn't certain about what the denominator of the ratio is.

  - The original experiment also uses stacked bar charts; this experiment only compares grouped bar charts.
With these in mind, I recommend that this paper not argue that this experiment is in any way a replication, but a completely new experiment inspired by previous work. This will require that the authors acknowledge the poor sampling techniques used as their own.

> "Partial" is in this case a reference to the fact that Cleveland & McGill's 1984 study was an extensive assessment of many different simple graphical elements; we are specifically focusing in this study on the simple element of comparisons on an aligned scale (e.g. grouped, but not stacked, bar charts). When extending C&M to 3D, it is necessary to focus on one simple element at a time, if only because of technological constraints - the technology exists (but is extremely expensive) to print multiple color objects that would be needed for stacked bar charts, and it is entirely unclear how one would 3D print a scatterplot. While we agree that there are not very many 3D-printed charts at this point in time, it does seem to be a promising area for visualization and making data tactile, but we have to lay the groundwork first. We hope to follow up with a 3D heatmap style plot, where the 3rd dimension is not redundant, but comparing to existing studies seemed wise.

- There are details missing in the description of the experiment.
Are the ratios participants are asked to estimate the same across the three
plot types (2D, 3D, and 3D printed)? Are all plots compared either sepa-
rated or not? How many times is a participant asked to estimate a ratio?
There are a lot of missing details here. I would recommend an ordered list
of tasks a participant is asked to do that emphasizes what variables (true
ratio, separated/not separated, graph dimension) are controlled or varied
at each step

> We have revised Figure 2 and the experiment description to clarify the design.

- The linear mixed effects model does not represent a thorough
statistical analysis of the data. This is true for several reasons. The
model seems very heavily parameterized for such few data. With my
confusion described in the previous point, it is not immediately clear to me
how many fixed effects there are, but it seems like there are 26 parameters
(two s.d.s from the random effects, 7 ratios, 2 graph types nested in 7
ratios (7*2), and 3 graph types) to fit using a small dataset. This is too
ambitious of a model. Furthermore, the normality assumptions on the
error term and the random effect are not assessed in any way. I would
recommend a simplier model with a proper assessment of the model fit.
Examining p-values is meaningless unless the model accurately emulates
the data generating mechanism.If a linear model seems inappropriate for these data (which seems likely
without a random sample), the authors may want to forsake p-values and
consider a more Machine Learning approach.

> We have revised the model and are now using a generalized additive model with spline terms for ratio effects fit by comparison type and display mode and a random effect for participant. While this model has more flexibility, it does have fewer parameters and effective degrees of freedom than the previous model, and examination of random effect terms and model residuals suggest that the model fits reasonably well.

- Please clearly state the hypotheses before calculating p-values.

> We do not include model fit statistics for hypothesis testing purposes, but as a mechanism to graphically examine the components of the variance. While p-values are technically provided in the model output, we primarily include these tables in order to provide information with which to interpret the accompanying figures.

- The paper implies a consensus from existing literature, but the
literature review presents several conflicting viewpoints. In the
abstract, the paper claims that numerous studies advocate to avoid 3D
visualizations when the third dimension does not convey any additional
information to the user, yet it is never explicitly stated which studies claim
this. The papers mentioned in the second paragraph of 1.1 comment on
accuracy and encoding information (with conflicting conclusions) and no
mention is given of any “consensus on 2D v. 3D.” This paper would benefit
greatly from a more focused literature review.

> We have improved the literature review; while the "3D is bad" advice is almost a trope in the graphics world, it is reasonable that people who are not visualization experts may not be familiar with this advice -- it is so common that it is even addressed by authors of packages meant for 3D graphics, such as rayshader.

- Figure 6 seems to imply a significant confounding variable that
is neither addressed nor investigated. It would appear that there is
a drop in the midmean of the log error when the true proportion is around
50%. With this observation, I find the argument that certain proportions
are easier to distinguish than others quite compelling. This should be
investigated. Furthermore, if this is true, the graph that a participant
randomly selects would have a strong effect on their ability to determine
a correct ratio.

> When we switched to using the generalized additive model, we spent some time investigating this effect. This is a fairly common result in perceptual experiments resulting from anchoring at endpoints and the midpoint of the line; it does not seem to result from the default slider value of 0.5. This is an interesting finding that merits a follow-up study.

- Abstract: As mentioned earlier, the ’numerous studies’ are not explicitedly listed anywhere.

> This has been addressed

- The acronym EPTs is introduced, but never used. Perhaps most blaring: the paper fails to use it immediately after introducing it (at the beginning of the next sentence).

> This has been addressed

- What is Mechanical Turk? Please provide a short explanation.

> Added a short explanation

- Relative language such as “often” and “largely” detract from the point the paper is trying to make, especially when this point does not seem to be backed up in the literature review. For instance, each of these words appear in the first sentence of Section 1.2.

> Some of this relative language is because it is difficult to make absolute statements about charts - certainly some people create charts in 3D (as we have in this study), but the vast majority of visualizations are intended for display on flat, 2D screens or papers. Similarly, we cannot guarantee that perceptual heuristics will always be applied in certain cases - for instance, people who have extreme amblyopia or have been blind in one eye since birth are immune to the line-width illusion, while those with binocular vision cannot overcome it. There are very few absolutes in perceptual research. We have attempted to clarify the language, but it is very difficult to remove all relative language here.

- Also on the first sentence of 1.2: I have failed to understand this point. What is the visual system that is “assuming” something? Is it a computer? Our eyes? Is there evidence that chart perception is so affected by this assumption?

> We have cited papers indicating that chart perception is indeed affected by the human visual system's heuristics. We have also added language clarifying that we are talking about the human perceptual system. In addition, addressing the line-width illusion using an example allows us to directly provide evidence to the reader that these effects are real.

- Please give a short explanation of the line-width illusion.

> Added.

Do the papers at the end of the first paragraph discuss the specific idea that 3D perceptions affect perceptions of 2D objects? Or do they just talk about these specific illusions? If it’s the former: move them up in this paragraph. The beginning of this paragraph feels like something that the authors are saying without evidence, as written.

> I believe this has been addressed because we re-wrote the paragraph explaining the connection between the line-width illusion and 2D/3D perception. Vanderplas & Hofmann (2015) in JCGS makes the case for the connection between 3D perception and the line width illusion in a compelling way; there is also an unpublished case study involving someone without binocular depth perception presented in Chapter 3.3 of https://github.com/srvanderplas/Dissertation/blob/master/thesis.pdf, if the reviewer is interested.

- second paragraph, first sentence: Did the subjects select these graphs for use, or did they better understand them? Are subjects here the people comprehending the graphs, or creating them? This is not clearly written.

> These short summaries of papers are all in the context of user studies, so these are designed experiments suggesting that participants prefer 2D charts for extracting information (that is, participants are presented with charts; they are not creating them).

- second paragraph: Why are the experienced subjects referred to as ’managers?’

> This study was evaluating usage of charts in a corporate setting - the experienced participants were literally managers within the company.

- second paragraph, second-to-last sentence, ’there are times where 3D graphs may better...’: Are these the same ’times’ when 2D graphs perform better? The two parts of this sentence could be interpreted as disparate ideas.

> As with most graphics research, there are always papers with contradictory results. We have tried to clarify the overall tone of this paragraph to assist with any confusion this may cause, but human subjects research is inherently messy.

- 1.2, second paragraph, last sentence: ’its’ should be ’their.’

> I believe this has been resolved

- 1.2, third paragraph, third sentence: ’has’ should be ’have’, unless ’Digital Graphics’ is the proper name for a field. In which case, capitalize ’graphics.’

> Fixed, thanks

- 1.2: Is this paper really ’reconsidering’ 3D charts? It does not sound like the literature really put them down.

> We've done a better job at supporting the idea that 3D graphics have been denigrated in visualization literature for at least the last 70 years; they were blamed for misleading graphics in "How to Lie with Statistics" in 1954, and I would suspect there are older snide comments, but they get harder to find in the pre-digital era. While 3D charts may be acceptable in many fields, in graphics and visualization they are very frequently condemned, mocked, and denigrated, perhaps without sufficient evidence.

- 1.2, ’these charts provide the opportunity...’: Make it clear that this is in reference to the 3D printed charts specifically.

> Fixed

- 1.3: Briefly discuss Cleveland and McGill’s theory, or do not mention it.

> Cleveland & McGill's papers produced a hierarchy of simple graphical elements, and while they also produced various theories, in this case, we're not actually citing any of their theories. In any case, I believe the language that triggered this comment has been removed in the edits.

- Figure 2: What’s the difference between Type 1 or Type 3?

> This has been clarified in the introduction, and we have used adjacent and separated as terms throughout this paper because they are a bit more intuitive.

- Discuss, or at least mention, Figures 2 and 3 in the text.

> Done

- 2.2, last paragraph, first sentence: What is covered in these papers? Are these two methods for digital rendering techniques? How are they different (and why does one require ’integration’ into another)?

> These are software packages used in this work - Shiny creates the web applet used to collect data, WebGL provides the interactive 3D renderings to allow us to show 3D plots within the browser.

- 2.3: Is the ratio the same across graph type? How do we know that participants didn’t use information from another plot type to drive their answers?

> Yes, the ratio is the same. As this was a relatively informal sample, we did check with participants verbally to see if they had any comments or had observed any interesting design quirks; most had no idea that the ratios were repeated. If statisticians didn't notice, we are fairly confident that their roommates who are not statisticians were also not aware. Examining the data, in 3/164 cases of (participant x ratio) trials participants had 0 variance on the provided estimates; in every case, these trials were ratios adjacent to 0.5 where participants did not adjust the slider. While this doesn't completely confirm that participants *couldnt* use information from other trials, it suggests that it was not a significant factor.

- 3.2, first paragraph, ”(through given the midmean...”: As written, this sounds like you are not expecting to find any log absolute error plots.

> This language is no longer present

- 3.2: Give more details on this Heer & Bostock model. Is there president for using a linear mixed effects model in this case.

> We've moved away from the LME towards a gam model (for which there is no precedent); if we could get the data from Heer & Bostock, we could examine differences between their model/data and ours, but that data is no longer available and we have reached out by email but have gotten no response.

- 4, first sentence: I don’t think that your literature review really suggested this.

> Addressed earlier

- 4, first paragraph, last sentence: What discrepancy? The one between you in the literature? If so, you really need to argue a stronger consensus in the literature, if there is one.

> Addressed earlier

- 4: I wouldn’t highlight that this paper does significance testing without more careful modeling.

> Addressed with better modeling and a bit more caution

- 4: “measurement method” should be “the measurement method.”

> Fixed

## Reviewer 2 Suggestions

> Thank you so much for your careful comments and suggestions - they have greatly improved the paper and we appreciate the thought you put into this.

There are a few edits that I view as necessary to strengthen the paper. These edits will help clarify important
issues for the reader, and also help avoid pitfalls.

-  I recommend adding a brief explanation and refernce for the line-width illusion in section 1.2. This would help readers unfamiliar with graphics research.

> We've added a figure demonstrating the sine illusion (as the simplest case of the line width illusion) and discussed in more depth the connection between this illusion and 3D graphics.

- At the end of section 1.2, it would be useful to add references, if there are any, for the comments of accessibility.

> We've added a paper by Chancey Fleet, who is blind and runs the NY Public Library lab for making visualizations available to the visually impaired. Hopefully that is sufficient.

- Section 2.3 could be clarified. Undertstanding the exact design with only the words in section 2.3 was difficult. Figure 2 helps, but was not cited in the text. Please clarify this a bit. This will also help the reader understand the model later, since the design is linked with the G(R) inclusion rather than having main effects and then an interaction.

> We've added a diagram which we hope will make the design much clearer. In addition, we've simplified the model so that hopefully it is easier to understand how the model works relative to the design.

- In Figure 6 the default smoother in ggplot2 was used. This accentuates the dip around x = 0.5, and allows the smallest and largest x values to have good bit of pull on the smoother. Since these features don’t seem to be important to the narrative, I recommend an adjustment be made.

> We have adjusted the figure to use the underlying data instead of the midmeans; this ensures that the effect is not just due to the default smoother (though we understand where that perception came from). In the process, we decided to use a gam model instead of the linear mixed effects model to account for the clear nonlinearity of the ratio relationship to the error. This has made the paper much more interesting (in our opinion), so thank you for calling this out, because it led to a thorough dive into the data that was quite rewarding.

- Pages 7-8. I recommend avoiding the phrase “testing for significant differences”, since this could be read incorrectly as not testing hypotheses set a priori (by someone rushing).

> I can't find that phrase anymore, but I hope that we've been sufficiently clear in our description that we're using the model in an informative/exploratory context more than as a means to do hypothesis testing.

- Page 8, bullet 1, I recommend acnowledging the full subscript in words

> I'm not sure what this is referring to, as there are no bullets on Page 8 that I can see.

- Details are needed as to what tests you ran from the LME. Only reporting p-values is not informative enough.

> Fair, but no longer relevant with the gam model. We've included model tables for informational purposes, but we're not really using them for significance testing (which is just as well since very few effects were significant, and even those effects need follow-up studies to ensure they aren't just an artifact of the experimental design and model-fitting process).

- In section 4, I found myself wondering if you would tell whether a participant interacted with a plot in the
shiny app. For example, could you track whether they clicked or dragged?

> Yes, we can, and we've added a section about this. We were surprised by how few people did actually interact with the 3D rendering; this may be a case where we need to add a 3D render to the tutorial so that people can learn how to play with it before the study. Thanks for the suggestion! In more recent studies, we've actually taken to returning the rotation matrix on each interaction and we're looking forward to trying to analyze that data!

- Section 1.1, paragraph 2: “that 2D graphs” should read “than 2D graphs”

> Fixed

- Page 2, paragraph 2, “Digital graphics has” should be “Digital graphics have”

> Fixed

- In section 1.3, I would continue to use the Cleveland and McGill (1984) citation style. That’s a stylistic preference, however.

> Fixed

- Page 4, paragraph 1. The citations Murdoch and Adler (2023) and Change at al. (2023) should be changed to rgl and shiny with parenthetical citations.

> Fixed

- Page 4, end of section 2.4. Change “3” to “Figure 3”.

> Fixed

- Figures 4 and 5 were somewhat hard to read. Could these be clearer?

> If these are the app screenshots, I am not sure how to make them clearer while still meeting the image resolution requirements for the journal, but we have tried.

- Section 3, how many cases were discarded? A sentence or two would avoid readers asking this question
and getting distracted.

> Fixed, thanks. Good catch.

-  Page 9, first line “impact of measurement” could be “impact of the measurement”

> Fixed

### 2024-01-22 jds.rmd
point-by-point response

## Elementary Graphical Tasks
@clevelandGraphical1984 established the comparative accuracy of different "elementary perceptual tasks" (EPTs).
Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
As a result of the Cleveland and McGill [-@clevelandGraphical1984] study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle. 
@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @clevelandGraphical1984 in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.

@@

## 3D Graphical Perception
Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion[@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013], for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in @fig:sine-illusion). 
While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception. Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.

@@

The use of 3D graphics have been explored in multiple studies. 
@fisher_data_1997 explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information. 
@barfield_effects_1989 compared 2D and 3D graphs presented on paper and on computers, showing that the accuracy of subject answers depended on their skill level: novice subjects were more accurate with 2D paper graphs, while experienced managers were more accurate with 3D computer graphs. 
For both experience levels, participants were more confident in their answers when using 2D graphs. 
There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may provide a more natural way to encode information, at least in theory. For instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately. 
@brath_3d_2014 highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data [@vanderplasTestingStatisticalCharts2020;@wilkeDonGo3D2020;@morganwallIntroducing3DGgplots20191006]. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Digital graphics have developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

In this paper, we discuss a study designed to examine Cleveland & McGill's experiments on grouped bar charts using modern graphics in two and three dimensions. This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data. In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information. <!-- This is included because of a graphics paper section comment that really annoyed me -->
In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from @clevelandGraphics1984 in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs. The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.

@@

@@

```{r midmeans-log-errors, fig.cap='Midmeans and observed values of log absolute errors for the true ratio of bars. Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='80%'}
plot.types = c('2D', 'Rendered 3D', '3D Printed')
names(plot.types) = c('2dDigital', '3dDigital', '3dPrint')

@@

## Linear Mixed Effects Model

In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a linear mixed effects model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio. This allows us to test for significant differences (though given the midmean log absolute error plots, we do not expect to find any) as well as to quantify effect sizes for future studies. The formal statistical model is as follows:

$$y_{ijklm}=\mu+S_i + R_j + G_k +T_l + \epsilon_{ijklm}$$

@@

- $S_i\sim N(0,\sigma^2_S)$ is the random effect of the $i^{th}$ subject

- $R_j$ is the effect of the $j^{th}$ ratio

@@

```{r, include=T}
library(papaja)
library(lme4)

mod <- lmer(response ~ (1|subject) + ratio + plot + type,
            data = res)
modsum <- summary(mod)
modaov <- anova(mod)
modaovsum <- summary(modaov)

@@

No differences were detected for the true ratio of bars ($t = `r modsum$coefficients[2,3]`$), whether the bars were adjacent or separated ($t = `r modsum$coefficients[5,3]`$), or for the plot nested within the true ratio ($F(`r modaov$npar[2]`, `r f_df`) = `r modaov[2,4]`$). 
While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes.

>>>>

## Elementary Graphical Tasks
@clevelandGraphical1984 established the comparative accuracy of different "elementary perceptual tasks" (EPTs).
Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
As a result of the Cleveland and McGill [-@clevelandGraphical1984] study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle. 
@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @clevelandGraphical1984 in an online setting using  Mechanical Turk, a platform for crowdsourcing human tasks that has been used for (among other things) assembling machine learning datasets. This replication study largely validated the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.

@@

## 3D Graphical Perception
Chart perception is affected by the human visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial. 
In addition, these artificial stimuli are typically created and presented in two dimensions.
The artificial scenes we create for data visualization have been documented to cause problems: for instance, the line-width illusion [@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013] has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, hammock plots, and Sankey diagrams. The line width illusion causes lines or error bands which are actually the same length to appear shorter at points with large amounts of curvature; a simple example of the classic sine illusion (one version of the line width illusion) is shown in \Cref{fig:sine-illusion}. 
While this illusion is problematic in two dimensions, when the depth cues are actually present in a three-dimensional situation, the same implicit visual heuristics contribute to accurate size perception (as in the middle panel of \Cref{fig:sine-illusion}). 
Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the visual display relative to the training of the visual system and the heuristics which are activated in order to make sense of the display.

@@

The use of 3D graphics have been explored in multiple studies, with mixed results. 
@fisher_data_1997 explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information. 
@barfield_effects_1989 compared 2D and 3D graphs presented on paper and on computers, showing that the accuracy of subject answers depended on their skill level: novice subjects were more accurate with 2D paper graphs, while experienced managers were more accurate with 3D computer graphs. 
For both experience levels, participants were more confident in their answers when using 2D graphs. 
There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may provide a more natural way to encode information, at least in theory. For instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately. 
@brath_3d_2014 highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.
Nevertheless, it is extremely common in visualization literature to recommend that 3D plots be avoided at all costs [@vanderplasTestingStatisticalCharts2020;@wilkeDonGo3D2020;@morganwallIntroducing3DGgplots20191006].

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data [@huffHowLieStatistics1954]. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Computer graphics and rendering technology have developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

In this paper, we discuss a study designed to examine Cleveland & McGill's experiments on grouped bar charts using modern graphics in two and three dimensions. This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data. In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information. <!-- This is included because of a graphics paper section comment that really annoyed me -->
In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from @clevelandGraphical1984 in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs. The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.

@@

```{r, include = F}
res %>%
  group_by(subject, ratio) %>%
  summarize(mean = mean(byHowMuch), sd = sd(byHowMuch))

@@

```{r midmeans-log-errors, fig.cap='Midmeans and observed values of log absolute errors for the true ratio of bars. Summary lines are computed from raw data using a loess smooth.', message=FALSE, warning=FALSE, out.width='\\textwidth'}
plot.types = c('2D', 'Rendered 3D', '3D Printed')
names(plot.types) = c('2dDigital', '3dDigital', '3dPrint')

@@

## Generalized Additive Model

In addition to replicating the (primarily graphical) analysis of participant errors, we also took a more statistical approach and fitted a generalized additive model that accounts for participant variation as well as the effect of comparison type, graph type, and ratio. The formal statistical model is as follows:

$$y_{ijklm}= s(R_j * T_l) + s(R_j *G_k) + T_l + G_k + s(S_i) + \epsilon_{ijklm}$$

@@

- $s(S_i)$ is the random effect of the $i^{th}$ subject

- $s(R_j*T_l)$ is the effect of the $j^{th}$ ratio, by comparison type

- $s(R_j*G_k)$ is the effect of the $j^{th}$ ratio, by display type

@@

```{r, include=F}
library(papaja)
library(lme4)

mod <- lmer(response ~ (1|subject) + ratioLabel + plot + type,
            data = res)
modsum <- summary(mod)
modaov <- anova(mod)
modaovsum <- summary(modaov)

@@

<!-- No differences were detected for the true ratio of bars ($t = `r modsum$coefficients[2,3]`$), whether the bars were adjacent or separated ($t = `r modsum$coefficients[5,3]`$), or for the plot nested within the true ratio ($F(`r modaov$npar[2]`, `r f_df`) = `r modaov[2,4]`$).  -->
<!-- The form of this linear mixed effects model assumes a linear relationship between ratio and response, but we know from \Cref{fig:midmeans-log-errors} that the actual relationship appears to be nonlinear.  -->
A generalized additive model with random effects for subject was fit using the `mgcv` package [@mgcv1;@mgcv2;@mgcv3]. Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}. It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in @hollandsBiasProportionJudgments2000.

While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` trials (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562). This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5. In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types. It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum. A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points  could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.

```{r gam}
library(mgcv)
library(gratia)
res2 <- res %>%
  mutate(participant = factor(as.numeric(factor(subject)))) %>%
  mutate(ratio_prop = ratioLabel/100) %>%
  mutate(type = factor(type, levels = 1:2, labels = c("Separated", "Adjacent"))) %>%
  mutate(plot = factor(plot)) %>%
  mutate(plottype = interaction(plot, type))
# res2

gam_mod <- gam(response ~ s(ratio_prop, by = plot, k = 6) + s(ratio_prop, by = type, k = 6) +  plot + type + s(participant, bs = "re"),
               data = res2, method = 'REML')
# gam_mod

gam_sum <- summary(gam_mod)
```

```{r, results = 'asis'}
gam_sum$p.table %>% knitr::kable(digits = 4, caption = "Parametric coefficients in gam model.", label = "gam-param", booktabs = T)

gam_sum$s.table %>% as.data.frame() %>% 
  mutate(Smooth = rownames(.),
         Smooth = str_replace(Smooth, "ratio_prop", "Ratio")) %>%
  select(Smooth, everything()) %>%
  knitr::kable(digits = 4, row.names = F, caption = "Approximate significance of smooth terms in gam model.", label = "gam-smooth", booktabs = T)
```

```{r gam-pred-chart, fig.cap = "Predictions with standard errors for fixed effects in a generalized additive model with splines for ratio by comparison and ratio by display method. It is clear that the separated comparisons are easier to estimate when small, right around 50\\%, or large, and harder to estimate between these points. What is interesting is that no such trend is present for adjacent comparisons. ", fig.width = 7, fig.height = 3, out.width = "\\textwidth"}
# 
# coef(gam_mod)
# draw(gam_mod, ncol = 4)

newdata <- expand.grid(ratio_prop = seq(0.15, .85, .001), type = c("Separated", "Adjacent"), plot = unique(res2$plot), participant = 1) %>%
  mutate(plottype = interaction(plot, type))
pred <- predict(gam_mod, newdata = newdata, type = "terms", se.fit = T) %>% as.data.frame()

pred_se <- select(pred, matches("se.fit")) %>%
  select(-matches("participant"))
pred_fit <- select(pred, matches("^fit.")) %>%
  select(-matches("participant"))

newdata %>%
  mutate(pred = rowSums(pred_fit), 
         se = rowSums(pred_se)) %>%
  ggplot(aes(x = ratio_prop, y = pred + gam_mod$coefficients[1], color = type)) + 
  geom_ribbon(aes(ymin = pred + gam_mod$coefficients[1] - 1.96*se, ymax = pred+ gam_mod$coefficients[1] + 1.96*se, fill = type), alpha = .25) +  
  geom_line() + facet_grid(.~plot) + 
  scale_color_discrete("Comparison", na.translate = F) + scale_fill_discrete("Comparison", na.translate = F) +

  geom_point(data = res, aes(x = ratioLabel/100, y = response, color = factor(type, levels = 1:2, labels = c("Separated", "Adjacent")))) +
  theme_bw() + ylab("Prediction") + xlab("True Proportion Difference") + 
  xlim(c(0, 1))

While \citeauthor{heerCrowdsourcingGraphicalPerception2010b} provided a zip file containing data and code for their paper, the link is no longer active, so it is not possible to fit a similar model to their data for comparison purposes; if we could get access to their data, this would allow for investigating whether the differences identified in our model were an artifact of the measurement method (e.g. slider vs. numerical entry) or if this effect was present in previous studies and can now be identified because of more advanced statistical modeling techniques.
